---
title: "COVID"
output: pdf_document
date: "2024-07-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

Reading in the various covid data files

```{r readdata, echo=TRUE}
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

file_names <- c("time_series_covid19_confirmed_US.csv",  "time_series_covid19_confirmed_global.csv", "time_series_covid19_deaths_US.csv",  "time_series_covid19_deaths_global.csv")

urls <- str_c(url_in,file_names)

confirmed_US <- read_csv(urls[1])
confirmed_global <- read_csv(urls[2])
deaths_US <- read_csv(urls[3])
deaths_global <- read_csv(urls[4])

```

### Data Summary
The data contains information on cases and deaths related to COVID-19 from both the US and Globally. There are many questions that can be answered with this data and we will complete multiple visual analysis to do so, particularly in predicting how well the virus is being responded to throughout the US.

### Data manipulations/clean ups
Below I am completing some data manipulations and clean ups in order to increase the quality of the data for further analysis

cleaning up the confirmed_global file
```{r cleanupconfirmedglobal, echo=TRUE}
confirmed_global <- confirmed_global %>%
    pivot_longer(cols = -c('Province/State', 'Country/Region', 'Lat', 'Long'), names_to = "date", values_to = "cases") %>%
    select(-c('Lat','Long'))
```

cleaning up the confirmed_US file
```{r cleanupconfirmedUS, echo=TRUE}
confirmed_US <- confirmed_US %>%
    pivot_longer(cols = -c('UID':'Combined_Key'), names_to = "date", values_to = "cases") %>%
    select(Admin2:cases) %>%
    mutate(date = mdy(date)) %>%
    select(-c(Lat,Long_)) 
```

cleaning up the deaths_US file
```{r cleanupdeathsUS, echo=TRUE}
deaths_US <- deaths_US %>%
    pivot_longer(cols = -(UID:Population), names_to = "date", values_to = "deaths") %>%
    select(Admin2:deaths)%>%
    mutate(date=mdy(date)) %>%
    select(-c(Lat, Long_))
```

cleaning up deaths_global file
```{r cleanupdeathsglobal, echo = TRUE}
deaths_global <- deaths_global %>%
    pivot_longer(cols = -c('Province/State', 'Country/Region', 'Lat', 'Long'), names_to = "date", values_to = "deaths") %>%
    select(-c('Lat','Long'))
```

join global cases and global deaths
```{r joinglobal, echo = TRUE}
global <- confirmed_global %>%
    full_join(deaths_global) %>%
    rename(Country_Region = 'Country/Region',
           Province_State = 'Province/State') %>%
    mutate(date = mdy(date))
```
Filter where cases global positive
```{r filtercases, echo=TRUE}
global <- global %>% filter(cases > 0)
```

Create Key in global
```{r globalkey, echo=TRUE}
global <- global %>%
        unite("Combined_Key", c(Province_State, Country_Region),
              sep=", ",
              na.rm = TRUE,
              remove = FALSE)
```

Join US cases and deaths
```{r USJoin, echo=TRUE}
US <- confirmed_US %>% 
      full_join(deaths_US)
```
US Analysis grouping
```{r groupingUSanalysis, echo=TRUE}
US_by_State <- US %>%
  group_by(Province_State, Country_Region, date) %>%
  summarize(cases=sum(cases), deaths=sum(deaths), Population=sum(Population)) %>%
  mutate(deaths_per_milly = deaths * 1000000 / Population) %>%
  mutate(cases_per_milly = cases * 1000000 / Population) %>%
  select(Province_State, Country_Region, date, cases, deaths, deaths_per_milly, cases_per_milly, Population) %>%
  ungroup()

US_total <- US %>%
  group_by(Country_Region, date) %>%
  summarize(cases=sum(cases), deaths=sum(deaths), Population=sum(Population)) %>%
  mutate(deaths_per_milly = deaths * 1000000 / Population) %>%
  mutate(cases_per_milly = cases * 1000000 / Population) %>%
  select(Country_Region, date, cases, deaths, deaths_per_milly, Population, cases_per_milly) %>%
  ungroup()
```

### US Visual
Looking at deaths and cases in the US at an aggregate level
```{r USanalysisVisual, echo=TRUE}
US_total %>%
  filter(cases>0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(colour = "cases")) +
  geom_point(aes(colour= "cases")) +
  geom_line(aes(y = deaths, colour = "deaths")) +
  geom_point(aes(y = deaths, colour = "deaths"))  + 
  scale_y_log10() +
  theme(legend.position = "bottom", axis.text.x = element_text(angle=90)) +
  labs(title = "COVID 19 in US", y=NULL)
```

### State Visual
Looking at deaths and cases in the state of New York at an aggregate level
```{r StateAnalysisVisual, echo=TRUE}
state <- "New York"
US_by_State %>%
  filter(Province_State == state) %>%
  filter(cases>0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(colour = "cases")) +
  geom_point(aes(colour= "cases")) +
  geom_line(aes(y = deaths, colour = "deaths")) +
  geom_point(aes(y = deaths, colour = "deaths"))  + 
  scale_y_log10() +
  theme(legend.position = "bottom", axis.text.x = element_text(angle=90)) +
  labs(title = str_c("COVID 19 in ",state), y=NULL)
  
```


### Analysis of new cases/deaths
Completing an analysis to understand if new cases/deaths are slowing down. This helps to see if the situation is getting better or not.

```{r ApplyLag, echo=TRUE}
US_by_State <- US_by_State %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))
US_total <- US_total %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))
```


```{r NewCasesDeathsTotal, echo=TRUE}
US_total %>%
  filter(cases>0) %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(colour = "new_cases")) +
  geom_point(aes(colour= "new_cases")) +
  geom_line(aes(y = new_deaths, colour = "new deaths")) +
  geom_point(aes(y = new_deaths, colour = "new deaths"))  + 
  scale_y_log10() +
  theme(legend.position = "bottom", axis.text.x = element_text(angle=90)) +
  labs(title = "New Cases/Deaths COVID 19 in US", y=NULL)
```
### New Cases Summary
It appears as though new cases have begun to flatten out, however there is clearly a large amount of variation in the later stages of the pandemic. I suspect this may be due to the quality of the data as monitoring becomes less important.

### New York Analysis
Applying visualization and tabular analysis of new cases/deaths to state to get a general idea of the situation and look for further areas to focus on.
```{r NewCasesDeathsState, echo=TRUE}
state <- "New York"
US_by_State %>%
  filter(Province_State == state) %>%
  filter(cases>0) %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(colour = "new cases")) +
  geom_point(aes(colour= "new cases")) +
  geom_line(aes(y = new_deaths, colour = "new deaths")) +
  geom_point(aes(y = new_deaths, colour = "new deaths"))  + 
  scale_y_log10() +
  theme(legend.position = "bottom", axis.text.x = element_text(angle=90)) +
  labs(title = str_c("COVID 19 in ",state), y=NULL)
```

```{r StateAnalysis, echo=TRUE}
US_state_totals <- US_by_State %>%
  group_by(Province_State) %>%
  summarize(deaths = max(deaths), cases = max(cases),
  population = max(Population),
  cases_per_thou = 1000* cases/population,
  deaths_per_thou = 1000 * deaths/population) %>%
  filter(cases>0, population>0)
```

```{r slicer, echo=TRUE}
US_state_totals %>%
  slice_min(deaths_per_thou, n=10) %>%
  select(deaths_per_thou, cases_per_thou, everything())
US_state_totals %>%
  slice_max(deaths_per_thou, n=10) %>%
  select(deaths_per_thou, cases_per_thou, everything())
```

### Linear Model
In this linear analysis I am analysing to see whether there is a linear correlation between deaths and cases. A further analysis on this data would be to segment the data into different time periods. This would help us understand if deaths per case were more correlated in the earlier stages of the pandemic compared to later stages (due to increased knowledge on virus etc.)
```{r linear, echo=TRUE}
mod= lm(deaths_per_thou ~ cases_per_thou, data=US_state_totals)
summary(mod)
US_tot_w_pred <- US_state_totals %>% mutate(pred=predict(mod))
US_tot_w_pred %>% ggplot() +
  geom_point(aes(x=cases_per_thou, y=deaths_per_thou), color = "blue") + geom_point(aes(x=cases_per_thou, y=pred), color ="red")
```

### Additional Visualization 1
visualization depicting whether the ratio of deaths per case was increasing or decreasing over time. e.g. Is treatment potentially improving or not.

```{r addtmodel, echo=TRUE}
US_by_State %>% mutate(difference = new_deaths/new_cases) %>%
filter(new_deaths >0, new_cases > 0, Province_State=="New York") %>%
  ggplot() + 
  geom_point(aes(x=date, y=difference), color="blue")
```

### Interpretation 1
Graph shows that early in the pandemic, there was a higher ratio of deaths per case. This makes sense and is likely due to a combination of uncertainty on how to deal with the novel virus as well as hospitals beign overwhelmed. There is then clear seasonality each year showing a worsening in the death to case ratio, this is likely again due to overwhelmed hospitals from other seasonal illness.

### Additional Visual Analysis 2
I am looking to create a cluster graph broken down by state showing the number of cases/deaths grouped by month. I am doing this analysis as I would like to see if there are distinct groups between states and to understand if states are consistent in their performance. This will help me to understand visually if there were better/worse responses between states. 


```{r addtmodel2, echo=TRUE}
US_by_State_Month <- US_by_State
US_by_State_Month$Month_Year <- as.Date(format(US_by_State_Month$date, "%Y-%m-01") )
cases_grouped <- US_by_State_Month %>%
  group_by(Month_Year, Province_State) %>%
  summarise(sumcase = sum(cases_per_milly), 
            sumdeath = sum(deaths_per_milly)) %>%
  filter(sumcase>0, sumdeath>0, sumcase < Inf)

ggplot() +
  geom_point(data = cases_grouped,
             show.legend = FALSE,
             mapping = aes(x = sumcase, 
                                  y = sumdeath, 
                                  colour = Province_State))

```
### Interpretation Analysis 2

This graph has a large amount of noise involved. This is somewhat what I expected, as this shows that there is not really a distinct state that is doing better in treatment of cases with the exception of the bottom couple of states outlined. This could be analyzed further to understand what these states did differently in order to have fewer deaths per case. 

### Additional Model
I am using a linear trend model in this analysis. What I am looking to understand is whether at a 'high level' the number of deaths within the state of New York is leveling off, or continuing in a positive fashion. 


```{r additionalmodel3, echo=TRUE}
cases_grouped <- cases_grouped %>%
  filter(Province_State == "New York")

p <- ggplot(cases_grouped, aes(x = Month_Year, y = sumdeath)) +
  geom_point(color = "red") +
  geom_line(color = "red") +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(x = "Month and Year", y = "Sum of Deaths") +
  theme_minimal()
print(p)

```

### Interpretation Analysis 3
Focusing in on New York again and tracking deaths over time, there is a clear upward linear trend (excluding obvious outliers). This is based on the sum of deaths, so of course this trend will never decrease, however it is concerning that there is continued upward trajectory of deaths. When performing this analysis I was hopeful to see a reduction in the increase in deaths however this does not appear to have occurred. A further analysis could involve a larger number of states to identify if this trend is consistent across various states focusing in on only new deaths (todays deaths - yesterdays deaths) to help clarify the picture. 



### Potential Bias
There is a posibility for both under/over reporting COVID cases and/or deaths, but it is difficult to know in what direction this bias is in. For example, there was political incentives to under report in order to show a better result from COVID restrictions however there may also be incentives to over-report in order to justify further restrictions or be able to sell certain products

### Overall Conclusion
Across the multiple analysis completed, it is concerning to see that for the most part cases were continuing to increase over time. Cases did not appear to be increasing exponentially, which is positive, however both cases and deaths were still showing increases across almost all periods of time. It would be incredibly interesting to see this data to current date, however due to expense of collecting, I understand this data is likely unavailable to the public.
